---
title: "Assignment 2: Coding Basics"
author: "Humayra Rahman"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons/labs in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document. 
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Canvas.

7.  Initial here to acknowledge that you did not use AI at all in completing this assignment: ___


## Basics, Part 1

1.  Use R's `seq()` function to create a sequence of numbers from 100 to 333, increasing by threes. Assign this sequence a variable name.
```{r}
seq(100, 333)
hundred_sequence <- seq(100, 333,3)
hundred_sequence

seq(100, 333, 3) # from, to, by
```

2.  Compute the *mean* of this sequence, assigning this values its own variable name. 
mean(hundred_sequence) # Vector type: Numeric

3.  Compute the *standard deviation* (`sd()`) of this sequence, assigning this values its own variable name.  
seq_sd <- sd(hundred_sequence)

4.  Display the the mean minus the standard deviation as well as the mean plus the standard deviation.
mean(hundred_sequence)-seq_sd
mean(hundred_sequence)+seq_sd
5.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. hundred_sequence <- seq(from = 100, to = 333, by = 3)

#2. seq_mean <- mean(hundred_sequence)

#3. seq_sd <- sd(hundred_sequence)

#4.seq_mean - seq_sd; seq_mean + seq_sd



```

---

## Basics, Part 2
6.  Create three vectors, each with four components, consisting of (a) student names, (b) test scores, and (c) whether they are on scholarship or not (TRUE or FALSE).
student_names <- c("Anna", "Ben", "Cary", "Don") # Vector type: character
test_scores <- c(75, 88, 95, 80) # Vector type: numeric
scholarship <- c(FALSE, TRUE, TRUE, FALSE) # Vector type: logic

7.  Label each vector with a comment on what type of vector it is.


8.  Combine each of the vectors into a data frame. Assign the data frame an informative name.
student_data <- data.frame(
student_names,test_scores,
scholarship
)


9.  Label the columns of your data frame with informative titles.
student_data <- data.frame(
  Student_Name = student_names,
  Test_Score = test_scores,
  Scholarship_Status = scholarship
)

```{r basics 2}


```

10.  QUESTION: How is this data frame different from a matrix?

> Answer: 
A data frame can hold different types of data in different columns, like names, numbers, and TRUE/FALSE values all together. A matrix can only store one type of data at a time. It can be numeric, character, or logical, but not a mix of all three at once.

---

## Basics, Part 3

11. Create a function with one input. In this function, use `if`...`else` to evaluate the value of the input: if it is greater than 50, it returns the word "Pass"; otherwise it returns the word "Fail". 
pass_fail <- function(x) {
  if (x > 50) {
    return("Pass")
  } else {
    return("Fail")
  }
}

pass_fail(83)
pass_fail(35)

12. Create a second function that does the exact same thing as the previous one but uses `ifelse()` instead if `if`...`else `. 
pass_fail_ifelse <- function(x) {
  ifelse(x > 50, "Pass", "Fail")
}

pass_fail_ifelse(50)

13. Run both functions using the value 54 as the input
pass_fail(54);pass_fail_ifelse(54)

14. Run both functions using the **vector** of student test scores you created as the input. (Only one will work properly...)


pass_fail_ifelse(test_scores)

```{r basics 3}
#11. Create a function using if...else
pass_fail <- function(x) {
  if (x > 60) {
    return("Pass")
  } else {
    return("Fail")
  }
}

#12. Create a function using ifelse()
pass_fail_ifelse <- function(x) {
  ifelse(x > 60, "Pass", "Fail")
}
#13a. Run the first function with the value 54
pass_fail <- function(x) {
  if (x > 54) {
    return("Pass")
  } else {
    return("Fail")
  }
}

#13b. Run the second function with the value 54
pass_fail_ifelse <- function(x) 
  ifelse(x > 54, "Pass", "Fail")

#14a. Run the first function with the vector of test scores
  pass_fail <- function(test_scores) {
  if (x > 54) {
    return("Pass")
  } else {
    return("Fail")
  }
}

#14b. Run the second function with the vector of test scores
```
pass_fail_ifelse <- function(test_scores) {
  ifelse(x > 54, "Pass", "Fail")
  
15. QUESTION: Which option of `if`...`else` vs. `ifelse` worked? Why? (Hint: search the web for "R vectorization"; it's ok here if an AI response is presented in the search response.)

> Answer:
ifelse() worked. It worked because ifelse() is vectorized, meaning it checks each value in the vector one by one and returns a result for each value. The if ... else statement only works with a single value, not a vector, so it does not work properly when given multiple values at once.

**NOTE** Before knitting, you'll need to comment out the call to the function in Q14 that does not work. (A document can't knit if the code it contains causes an error!)