---
title: "Assignment 6: Data Scraping"
author: "Humayra Rahman"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 setting up packages
library(tidyverse)
library(rvest)
library(lubridate)
library(here)

getwd()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Set the year to 2024
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2
#scraping website
the_url <-
"https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024"
the_website <- read_html(the_url)

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * Contact Person
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "Reginald Hicks", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
#reading all tables on the page into a list
all_tables <- the_website %>%
  html_nodes("table") %>%
  html_table(fill = TRUE)

#water system name
water_system_name <- the_website %>%
  html_nodes("table:nth-of-type(1) tr:nth-child(1) td:nth-child(2)") %>%
  html_text() %>%
  .[1]
water_system_name

#contact person
contact_person <- the_website %>%
  html_nodes("table:nth-of-type(1) tr:nth-child(3) td:nth-child(2)") %>%
  html_text() %>%
  .[1]
contact_person 

#ownership
ownership <- the_website %>%
  html_nodes("table:nth-of-type(1) tr:nth-child(2) td:nth-child(4)") %>%
  html_text() %>%
  .[1]
ownership

#MGD for each month
monthly_table <- all_tables[[13]]

max_withdrawals <- c(
  monthly_table[[3]][1:4],  # Jan, Feb, Mar, Apr
  monthly_table[[6]][1:4],  # May, Jun, Jul, Aug
  monthly_table[[9]][1:4]   # Sep, Oct, Nov, Dec
)

max_withdrawals          
length(max_withdrawals) 


```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 

#month order according to scraping
month_order <- c(1,2,3,4, 5,6,7,8, 9,10,11,12)

df_withdrawals <- data.frame(
  "Month"          = month_order,
  "Year"           = rep(2024, 12),
  "Max_Use_mgd"    = as.numeric(max_withdrawals),
  "Water_System"   = rep(water_system_name, 12),
  "Contact_Person" = rep(contact_person, 12),
  "Ownership"      = rep(ownership, 12)
) %>%
  mutate(Date = my(paste(Month, "-", Year)))

glimpse(df_withdrawals)

#5

#creating a line plot
ggplot(df_withdrawals, aes(x = Date, y = Max_Use_mgd)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2.5) +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  labs(
    title = "2024 Max Daily Water Withdrawals - Durham",
    x     = "Month",
    y     = "Max Day Use (MGD)"
  ) +
  theme_classic()

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs, "PWSID" and "year", that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 

#constructing func
scrape.it <- function(PWSID, year) {
  
#creating URL from inputs
the_url <- paste0(
"https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=",
PWSID, "&year=", year
  )
  
#fetching page
the_website <- read_html(the_url)
  
#reading all tables
all_tables <- the_website %>%
  html_nodes("table") %>%
  html_table(fill = TRUE)
  
#scraping system info
water_system_name <- the_website %>%
  html_nodes("table:nth-of-type(1) tr:nth-child(1) td:nth-child(2)") %>%
    html_text() %>%
    .[1]
  
contact_person <- the_website %>%
  html_nodes("table:nth-of-type(1) tr:nth-child(4) td:nth-child(2)") %>%
    html_text() %>%
    .[1]
  
ownership <- the_website %>%
  html_nodes("table:nth-of-type(1) tr:nth-child(2) td:nth-child(4)") %>%
    html_text() %>%
    .[1]
  
#finding monthly table
monthly_index <- which(sapply(all_tables, function(t)
  ncol(t) == 9 & nrow(t) == 4))[1]
  monthly_table <- all_tables[[monthly_index]]
  
#scraping max dat use
  max_withdrawals <- c(
    monthly_table[[3]][1:4],  # Jan, Feb, Mar, Apr
    monthly_table[[6]][1:4],  # May, Jun, Jul, Aug
    monthly_table[[9]][1:4]   # Sep, Oct, Nov, Dec
  )
  
#building dataframe
df <- data.frame(
    "Month"          = 1:12,
    "Year"           = rep(year, 12),
    "Max_Use_mgd"    = as.numeric(max_withdrawals),
    "Water_System"   = rep(water_system_name, 12),
    "Contact_Person" = rep(contact_person, 12),
    "Ownership"      = rep(ownership, 12),
    "PWSID"          = rep(PWSID, 12)
  ) %>%
    mutate(Date = my(paste(Month, "-", Year)))
  
  
  return(df)
}



```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 

#extracting Durham 2020 data using the function
df_durham_2020 <- scrape.it(PWSID = "03-32-010", year = 2020)

#showing structure
glimpse(df_durham_2020)

#plotting
ggplot(df_durham_2020, aes(x = Date, y = Max_Use_mgd)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2.5) +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  labs(
    title = "2020 Max Daily Water Withdrawals - Durham",
    x     = "Month",
    y     = "Max Day Use (MGD)"
  ) +
  theme_classic()

```

8. Use the function above to extract data for Cary (PWSID = '03-92-020') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 

#extracting Cary 2020 data
df_cary_2020 <- scrape.it(PWSID = "03-92-020", year = 2020)

#combining Durham and Cary into one dataframe
df_combined_2020 <- bind_rows(df_durham_2020, df_cary_2020)

#comparison plot
ggplot(df_combined_2020, aes(x = Date, y = Max_Use_mgd, color = Water_System)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  labs(
    title = "2020 Max Daily Water Withdrawals: Durham vs. Cary",
    x     = "Month",
    y     = "Max Day Use (MGD)",
    color = "Water System"
  ) +
  theme_classic() +
  theme(legend.position = "top")

```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023. Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "06_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs (where one of the two inputs does not change). Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 

#creating vectors of inputs
the_years   <- 2018:2023
cary_pwsids <- rep("03-92-020", length(the_years))

#using map2
df_cary_multiyear <- map2(cary_pwsids, the_years, scrape.it) %>%
  bind_rows()

#plotting w loess smooth
ggplot(df_cary_multiyear, aes(x = Date, y = Max_Use_mgd)) +
  geom_line(color = "steelblue", linewidth = 0.8, alpha = 0.6) +
  geom_point(color = "steelblue", size = 1.5) +
  geom_smooth(method = "loess", se = FALSE,
              color = "darkred", linewidth = 1.2) +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  labs(
    title = "Cary Max Daily Water Withdrawals: 2018-2023",
    x     = "Year",
    y     = "Max Day Use (MGD)"
  ) +
  theme_classic()
```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer: Yes. The LOESS line rises gradually from ~22 MGD in 2018 to ~24 MGD by 2022, then levels off, suggesting modest growth in water demand that appears to be stabilizing in recent years.
>
